---
title: "1_plots_and_first_analyses_rapeseed"
date: "2025-04-08"
output: html_document
---

```{r Packages}
library(tidyverse)
library(readr)
library(vegan)
library(ggvenn)
library(EnvStats)
library(ggpubr)
library(scales)
library(ComplexUpset)
library(patchwork)
```

```{loading data}

otu_min_1000 <- read_csv("otu_min_1000.csv", col_types = cols(...1 = col_skip()))


meta <- read_csv("meta_final_final.csv")

ids_1000 <- read_csv("ids_1000.csv")


##pest_data


all_tab <- read_csv("pest_data.csv")


meta_traps<-read_csv("meta_traps.csv")

```{r raw specnum and adding to meta_final}

otu_trans_1000 <- otu_min_1000 %>%
pivot_longer(cols=c(-query),names_to="sample")%>%
pivot_wider(names_from=c(query))

raw_spec_num <- otu_trans_1000 %>% 
  column_to_rownames(., var = "sample") %>% 
  specnumber() %>% 
  enframe() %>% 
  rename(sample = name,
         sr = value)

meta_final <- meta_final %>% 
  left_join(.,raw_spec_num)

##pest species

raw_spec_num_pest <- all_tab %>% 
  column_to_rownames(., var = "sample") %>% 
  specnumber() %>% 
  enframe() %>% 
  rename(sample = name,
         sr = value)

meta_all <- meta_all %>% 
  left_join(.,raw_spec_num)


```
```{Venn plots with the types}


venn_otu_stem_1000 <- otu_trans_1000  %>%
left_join(meta_final, by = "sample") %>%
filter(type == "Stem") %>%
select(-"sample",-"type",-"site",-"trip", -"sr") %>%
colSums() %>%
as.data.frame() %>%
filter(. != "0") %>%
rownames_to_column("query") %>%
select(query) %>%
unlist() %>%
as.character()

venn_otu_wipe_1000 <- otu_trans_1000  %>%
left_join(meta_final, by = "sample") %>%
filter(type == "Wipe") %>%
select(-"sample",-"type",-"site",-"trip", -"sr") %>%
colSums() %>%
as.data.frame() %>%
filter(. != "0") %>%
rownames_to_column("query") %>%
select(query) %>%
unlist() %>%
as.character()

venn_otu_flower_1000 <- otu_trans_1000  %>%
left_join(meta_final, by = "sample") %>%
filter(type == "Flower") %>%
select(-"sample",-"type",-"site",-"trip", -"sr") %>%
colSums() %>%
as.data.frame() %>%
filter(. != "0") %>%
rownames_to_column("query") %>%
select(query) %>%
unlist() %>%
as.character()

venn_otu_ff_1000 <- otu_trans_1000  %>%
left_join(meta_final, by = "sample") %>%
filter(type == "Fake flower") %>%
select(-"sample",-"type",-"site",-"trip", -"sr") %>%
colSums() %>%
as.data.frame() %>%
filter(. != "0") %>%
rownames_to_column("query") %>%
select(query) %>%
unlist() %>%
as.character()


venn_data_methods_1000 <- list("Fake Flower" = venn_otu_ff_1000, "Flower" = venn_otu_flower_1000, "Stem" = venn_otu_stem_1000, "Wipe" = venn_otu_wipe_1000)

ggvenn(data = venn_data_methods_1000 , show_percentage = F, fill_color = c("orange","yellow","green","darkgrey"),fill_alpha = .5,text_size = 6, set_name_size = 5)

```

```{UpSet plot}

venn_otu_stem_pests <- all_tab  %>%
left_join(meta_all, by = "sample") %>%
filter(type == "Stem") %>%
select(-"sample",-"type",-"site",-"trip") %>%
colSums() %>%
as.data.frame() %>%
filter(. != "0") %>%
rownames_to_column("query") %>%
select(query) %>%
unlist() %>%
as.character()

venn_otu_ff_pests <- all_tab  %>%
left_join(meta_all, by = "sample") %>%
filter(type == "Fake flower") %>%
select(-"sample",-"type",-"site",-"trip") %>%
colSums() %>%
as.data.frame() %>%
filter(. != "0") %>%
rownames_to_column("query") %>%
select(query) %>%
unlist() %>%
as.character()

venn_otu_flower_pests <- all_tab  %>%
left_join(meta_all, by = "sample") %>%
filter(type == "Flower") %>%
select(-"sample",-"type",-"site",-"trip") %>%
colSums() %>%
as.data.frame() %>%
filter(. != "0") %>%
rownames_to_column("query") %>%
select(query) %>%
unlist() %>%
as.character()

venn_otu_yellow_pests <- all_tab  %>%
left_join(meta_all, by = "sample") %>%
filter(type == "Yellow") %>%
select(-"sample",-"type",-"site",-"trip") %>%
colSums() %>%
as.data.frame() %>%
filter(. != "0") %>%
rownames_to_column("query") %>%
select(query) %>%
unlist() %>%
as.character()

venn_otu_wipe_pests <- all_tab  %>%
left_join(meta_all, by = "sample") %>%
filter(type == "Wipe") %>%
select(-"sample",-"type",-"site",-"trip") %>%
colSums() %>%
as.data.frame() %>%
filter(. != "0") %>%
rownames_to_column("query") %>%
select(query) %>%
unlist() %>%
as.character()


venn_data_methods_pests <- list("Fake Flower" = venn_otu_ff_pests, "Flower" = venn_otu_flower_pests, "Stem" = venn_otu_stem_pests, "Wipe" = venn_otu_wipe_pests, "Yellow pan trap" = venn_otu_yellow_pests)

all_otus <- unique(unlist(venn_data_methods_pests))
sets <- names(venn_data_methods_pests)

upset_df <- data.frame(
row.names = all_otus,
sapply(sets, function(s) as.integer(all_otus %in% venn_data_methods_pests[[s]]))
)

###small change

upset_df$"Fake Flower"<-upset_df$Fake.Flower

upset_df$"Yellow pan trap"<-upset_df$Yellow.pan.trap

upset_df<-upset_df %>% select(-Yellow.pan.trap, -Fake.Flower)

###plot

upset(
upset_df,
sets,               # the column names to use as sets
min_size = 1
)


youwaupset<-upsetme + theme(text= element_text(size = 16),title = element_text(size = 16))


```

```{NMDS}

set.seed(8)

nmds_1000 <- otu_trans_1000 %>%  
  left_join(meta_final,by = "sample") %>% 
  select(sample, contains("otu")) %>% 
  column_to_rownames(., var = "sample") %>%
  vegan::metaMDS(distance = "jaccard", k = 2, binary = T, trymax = 300)

NMDSscores_1000 <- scores(nmds_1000, display = "sites") %>%
 as.data.frame() %>%
 rownames_to_column("sample") %>%
 left_join(meta_final,., by = "sample") %>%
 drop_na(NMDS1)%>% 
 drop_na(NMDS2)

###BY METHOD

nmds_type_1000 <- NMDSscores_1000  %>%
ggplot(aes(x = NMDS1, y = NMDS2, color = type, shape=site)) +
geom_point(size=3) +
stat_ellipse(aes(group = type), type = "norm", level = 0.95, geom = "path", alpha = 1, size = 1) +
labs(color="Method") +
theme_minimal() + annotate("text", x = Inf, y = -Inf, hjust = 1, vjust = 0,
label = paste("Stress:", round(nmds_1000$stress, 4)))+
theme(legend.text = element_text(size = 16),
legend.title = element_text(size = 20))+
theme(axis.text= element_text(size = 16),
axis.title = element_text(size = 20))

###pests

###removal of empty columns first

all_tabminus<- all_tab %>%
right_join(meta_trimmed_pests,by = "sample")

all_tabminus<-all_tabminus[,-c(17:20)]


nmds_pests <- all_tabminus %>%
column_to_rownames(., var = "sample") %>%
vegan::metaMDS(distance = "jaccard", k = 2, binary = T, trymax = 300)


NMDSscores_pests <- scores(nmds_pests, display = "sites") %>%
as.data.frame() %>%
rownames_to_column("sample") %>%
left_join(meta_trimmed_pests,., by = "sample") %>%
na.omit(NMDS1)


nmds_type_pests <- NMDSscores_pests  %>%
ggplot(aes(x = NMDS1, y = NMDS2, color = type, shape = site)) +
geom_point(size=3) +
stat_ellipse(aes(group = type), type = "norm", level = 0.95, geom = "path", alpha = 1, size = 1) +
labs(color="Method") +
theme_minimal() + annotate("text", x = Inf, y = -Inf, hjust = 1, vjust = 0,
label = paste("Stress:", round(nmds_pests$stress, 4)))+
theme(legend.text = element_text(size = 20),
legend.title = element_text(size = 16))+
theme(axis.text= element_text(size = 16),
axis.title = element_text(size = 20))



```

```{Mean plots}

meanplot_methods<-meta_final %>%
mutate(type=factor(type, levels = c("Fake flower", "Flower", "Stem", "Wipe" , "Yellow"))) %>%
arrange(type) %>%
ggerrorplot(x = "type", y = "sr",
desc_stat = "mean_sd",
error.plot = "errorbar",            # Change error plot type
add = "mean"                        # Add mean points
) +
theme(legend.text = element_text(size = 16),
legend.title = element_text(size = 20))+
theme(axis.text= element_text(size = 16),
axis.title = element_text(size = 20))+
labs(y = "OTU richness",
x = "Method") +
stat_n_text()

###pests

meanplot_ordered_pests<-meta_all %>%
mutate(type=factor(type, levels = c("Fake flower", "Flower", "Stem", "Wipe" , "Yellow"))) %>%
arrange(type) %>%
ggerrorplot(x = "type", y = "sr",
desc_stat = "mean_sd",
error.plot = "errorbar",            # Change error plot type
add = "mean"                        # Add mean points
) +
theme(legend.text = element_text(size = 16),
legend.title = element_text(size = 20))+
theme(axis.text= element_text(size = 16),
axis.title = element_text(size = 20))+
labs(y = "Pest richness",
x = "Method") +
stat_n_text()




```{bar chart}


tax_order <- ids_1000 %>% 
    mutate(order = case_when(str_detect(class, "Collembola") ~ "Collembola",
                             TRUE ~ as.character(order))) %>% 
    mutate(order = case_when(str_detect(subclass, "Acari") ~ "Acari",
                             TRUE ~ as.character(order))) %>%
    filter(perc_id >= 90) %>% 
    select(query, order)

###GETTING PROPORTION OF ALL TAXA
###EVALUATING OTU NUMBERS

 rel_otu_num<-otu_trans_1000 %>% 
     pivot_longer(cols = -sample, names_to = "query", values_to = "reads") %>% 
     inner_join(tax_order, by = "query") %>%
     drop_na(order) %>% 
     mutate(order = factor(order)) %>% 
     filter(reads > 0) %>%
     group_by(sample, order) %>% 
     summarise(n_queries = n_distinct(query), .groups = "drop") %>% 
     group_by(sample) %>% 
     mutate(proportion = n_queries / sum(n_queries)) %>% 
     ungroup()


```

safe_colorblind_palette <- c("#88CCEE", "#E69F00", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499",
"#44AA99", "#999933", "#882255", "#000000", "#888888")

###more colors needed?

safe_colorblind_palette_species_vibecheck <- c("#88CCEE", "#CC6677", "#DDCC77","#DDCD99" , "#117733", "#117760", "#332288", "#AA4499",
"#44AA99", "#999933", "#882255", "#000000", "#888888", "#6699CC", "#661100")


BIGGER_relplot<-rel_otu_num %>%
left_join(meta_final) %>%
na.omit() %>%
select(sample, order, type, proportion) %>%
group_by(type) %>%
mutate(sum_time = sum(proportion)) %>%
ungroup() %>%
group_by(type, order) %>%
mutate(rel_abund_time = sum(proportion)) %>%
ungroup() %>%
group_by(type,order) %>%
summarize(rel_abund_time = mean(rel_abund_time/sum_time)) %>%
mutate(order2 = case_when(rel_abund_time < .01 ~ "< 1% rel. number",
TRUE ~ as.character(order))) %>%
group_by(order2,type) %>% summarise(rel_abund_time  = sum(rel_abund_time)) %>%
ggplot() +
geom_col(aes(x = type, y = rel_abund_time, fill = order2), color = "black") +
scale_fill_manual(values = safe_colorblind_palette) +
theme_bw() +
scale_y_continuous(labels = scales::percent_format())+
xlab("Methods") +
ylab("Relative number of OTUs (%)") +
guides(fill=guide_legend(title="Taxa")) +
theme(legend.text = element_text(size = 18),
legend.title = element_text(size = 20))+
theme(axis.text= element_text(size = 18),
axis.title = element_text(size = 20))


###pests - OCCUPANCY BARPLOT

rel_abund_allpests<-all_tab %>% pivot_longer(cols = -sample, names_to = "query", values_to = "reads") %>%
mutate(order = factor(query)) %>%
group_by(sample, order) %>%
summarise(read_order = sum(reads)) %>%
ungroup() %>%
group_by(sample) %>%
mutate(read_sum = sum(read_order)) %>%
ungroup() %>%
group_by(sample, order) %>%
summarise(rel_abund  = read_order/read_sum) %>%
ungroup() %>%
mutate(order2 = case_when(rel_abund < .01 ~ "< 1% rel. abundance",
TRUE ~ as.character(order))) %>%
group_by(sample, order2) %>%
summarise(rel_abund  = sum(rel_abund))

a_pestall_method<-rel_abund_allpests %>%
left_join(meta_all) %>%
na.omit() %>%
select(sample, order2, type, rel_abund) %>%
group_by(type) %>%
mutate(sum_time = sum(rel_abund)) %>%
ungroup() %>%
group_by(type, order2) %>%
mutate(rel_abund_time = sum(rel_abund)) %>%
ungroup() %>%
group_by(type,order2) %>%
summarize(rel_abund_time = mean(rel_abund_time/sum_time)) %>%
mutate(order2 = case_when(rel_abund_time < .01 ~ "< 1% rel. abundance",
TRUE ~ as.character(order2))) %>%
mutate(order2 = case_when(order2 == "Nan" ~ "< 1% rel. abundance",
TRUE ~ as.character(order2))) %>%
group_by(order2,type) %>% summarise(rel_abund_time  = sum(rel_abund_time)) %>%
ggplot() +
geom_col(aes(x = site, y = rel_abund_time, fill = order2), color = "black") +
scale_fill_manual(values = safe_colorblind_palette_species_vibecheck) +
theme_bw() +
scale_y_continuous(labels = scales::percent_format())+
xlab("Method") +
ylab("Relative occupancy") +
guides(fill=guide_legend(title="Taxa (pest only)")) +
theme(legend.text = element_text(size = 16),
legend.title = element_text(size = 20))+
theme(axis.text= element_text(size = 16),
axis.title = element_text(size = 20))


```{Abundances species}


less_strict_tax_species_bota <- ids_1000 %>%
mutate(species = case_when((species=="Meligethes viridescens") ~ "Brassicogethes viridescens",
TRUE ~ as.character(species))) %>%
mutate(species = case_when((species=="Brassicogethes sp. O130") ~ "Brassicogethes aeneus",
TRUE ~ as.character(species))) %>%
mutate(species = case_when((species=="Ceutorhynchus assimilis") ~ "Ceutorhynchus obstrictus",
TRUE ~ as.character(species))) %>%
mutate(species = case_when((species=="Botanophila sp. O111") ~ "Botanophila fugax",
TRUE ~ as.character(species))) %>%
mutate(species = case_when(str_detect(species, "Smittia sp. BIOUG21999-A08") ~ "Bryophaenocladius terrestris",
TRUE ~ as.character(species))) %>%
mutate(species = case_when((species=="Smittia sp. 8ES") ~ "Bryophaenocladius terrestris",
TRUE ~ as.character(species)))
filter(perc_id >= 100) %>%
select(query, species)


less_strict_rel_abund_species_no1percent_bota <- otu_trans_1000 %>%
pivot_longer(cols = -sample, names_to = "query", values_to = "reads") %>%
left_join(less_strict_tax_species_bota , by = "query") %>%
drop_na(species) %>%
mutate(species = factor(species)) %>%
group_by(sample, species) %>%
summarise(read_species = sum(reads)) %>%
ungroup() %>%
group_by(sample) %>%
mutate(read_sum = sum(read_species)) %>%
ungroup() %>%
group_by(sample, species) %>%
summarise(rel_abund  = read_species/read_sum) %>%
ungroup() %>%
group_by(sample, species) %>%
summarise(rel_abund  = sum(rel_abund))


```{patchwork}

meanplot_methods + venn_3b + BIGGER_nmds_k2 + BIGGER_relplot + plot_annotation(tag_levels = c('A'))


(BIGGER_mean_pests + BIGGER_pest_nmds) /
(wrap_elements(full = youwaupset) + no1pc_occ_plot) + plot_annotation(tag_levels = "A")


```