---
title: "1_plots_and_first_analyses"
date: "2024-04-12"
output: html_document
---

```{r Packages}
library(tidyverse)
library(readr)
library(vegan)
library(ggvenn)
library(ggpubr)
library(EnvStats)
library(patchwork)
library(UpSetR)

```

```{loading data}

new_otu <- read_csv("new_otu.csv", col_types = cols(...1 = col_skip()))


meta <- read_csv("meta.csv", col_types = cols(...1 = col_skip()))


new_taxa <- read_csv("new_taxa.csv", col_types = cols(...1 = col_skip()))



```{r raw specnum and adding to meta}

raw_spec_num <- new_otu %>% 
  column_to_rownames(., var = "sample") %>% 
  specnumber() %>% 
  enframe() %>% 
  rename(sample = name,
         sr = value)

meta <- meta %>% 
  left_join(.,raw_spec_num)

```

```{Venn plots with the methods}

venn_otu_wipe <- new_otu %>% 
  inner_join(meta %>%  
  select ("sample","method"), by = "sample") %>% 
  filter(method == "Wipe") %>% 
  select(-"sample",-"method") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  filter(. != "0") %>% 
  rownames_to_column("otu_id") %>% 
  select(otu_id) %>% 
  unlist() %>% 
  as.character()

venn_otu_fake <- new_otu %>% 
  inner_join(meta %>%  
  select ("sample","method"), by = "sample") %>% 
  filter(method == "Fake flower") %>% 
  select(-"sample",-"method") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  filter(. != "0") %>% 
  rownames_to_column("otu_id") %>% 
  select(otu_id) %>% 
  unlist() %>% 
  as.character()

venn_otu_malaise <- new_otu %>% 
  inner_join(meta %>%  
  select ("sample","method"), by = "sample") %>% 
  filter(method == "Malaise") %>% 
  select(-"sample",-"method") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  filter(. != "0") %>% 
  rownames_to_column("otu_id") %>% 
  select(otu_id) %>% 
  unlist() %>% 
  as.character()

venn_otu_netting <- new_otu %>% 
  inner_join(meta %>%  
  select ("sample","method"), by = "sample") %>% 
  filter(method == "Netting") %>% 
  select(-"sample",-"method") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  filter(. != "0") %>% 
  rownames_to_column("otu_id") %>% 
  select(otu_id) %>% 
  unlist() %>% 
  as.character()

venn_otu_yellow <- new_otu %>% 
  inner_join(meta %>%  
  select ("sample","method"), by = "sample") %>% 
  filter(method == "Yellow") %>% 
  select(-"sample",-"method") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  filter(. != "0") %>% 
  rownames_to_column("otu_id") %>% 
  select(otu_id) %>% 
  unlist() %>% 
  as.character()

venn_otu_plant <- new_otu %>% 
  inner_join(meta %>%  
  select ("sample","method"), by = "sample") %>% 
  filter(method == "Plant") %>% 
  select(-"sample",-"method") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  filter(. != "0") %>% 
  rownames_to_column("otu_id") %>% 
  select(otu_id) %>% 
  unlist() %>% 
  as.character()

venn_data_fakenadyellow <- list("Fake Flower" = venn_otu_fake, "Yellow" = venn_otu_yellow)

venn_data_nwpa <- list("Netting" = venn_otu_netting, "Wipe" = venn_otu_wipe, "Plant" = venn_otu_plant)

forcowplot_fakeandyellow<-ggvenn(data = venn_data_fakenadyellow, show_percentage = F, fill_color = c("#E69F00","#F0E442"),fill_alpha = .5,text_size = 6, set_name_size = 6)

forcowplot_nwpa<-ggvenn(data = venn_data_nwpa , show_percentage = F, fill_color = c("#CC79A7","#999999","#999903"),fill_alpha = .5,text_size = 6, set_name_size = 6)

venn_otu_sugar <- new_otu %>% 
  inner_join(meta %>%  
  select ("sample","infos"), by = "sample") %>% 
  filter(infos == "with sugar") %>% 
  select(-"sample",-"infos") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  filter(. != "0") %>% 
  rownames_to_column("otu_id") %>% 
  select(otu_id) %>% 
  unlist() %>% 
  as.character()

venn_otu_no_sugar <- new_otu %>% 
  inner_join(meta %>%  
  select ("sample","infos"), by = "sample") %>% 
  filter(infos == "no sugar") %>% 
  select(-"sample",-"infos") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  filter(. != "0") %>% 
  rownames_to_column("otu_id") %>% 
  select(otu_id) %>% 
  unlist() %>% 
  as.character()

venn_data_sugarornot <- list("Fake Flower (soaked with sugar)" = venn_otu_sugar, "Fake Flower (no sugar)" = venn_otu_no_sugar, "Yellow Trap" = venn_otu_yellow)

ggvenn(data = venn_data_sugarornot, show_percentage = F, fill_color = c("red","orange","yellow"),fill_alpha = .5,text_size = 6, set_name_size = 5)


```

```{NMDS}

set.seed(8)

nmds <- new_otu %>%  
  left_join(meta,by = "sample") %>% 
  select(sample, contains("otu")) %>% 
  column_to_rownames(., var = "sample") %>% 
  vegan::metaMDS(distance = "jaccard", k = 2, binary = T, trymax = 999)

NMDSscores <- scores(nmds, display = "sites") %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  left_join(meta,., by = "sample")%>% 
  drop_na(NMDS1)%>% 
  drop_na(NMDS2)

THIS TO EDIT nmds_group <- NMDSscores  %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = group)) +
  geom_point(size=3) + 
  stat_ellipse(aes(group = group), type = "norm", level = 0.95, geom = "path", alpha = 1, size = 1) +
  theme_minimal() + annotate("text", x = Inf, y = -Inf, hjust = 1, vjust = 0,
           label = paste("Stress:", round(nmds$stress, 4)))+
theme(legend.text = element_text(size = 16),
legend.title = element_text(size = 20))+
theme(axis.text= element_text(size = 16),
axis.title = element_text(size = 20))

nmds_method <- NMDSscores  %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = method, shape=infos)) +
  geom_point(size=3) + 
  stat_ellipse(aes(group = method), type = "norm", level = 0.95, geom = "path", alpha = 1, size = 1) +
  theme_minimal() + annotate("text", x = Inf, y = -Inf, hjust = 1, vjust = 0,
           label = paste("Stress:", round(nmds$stress, 4)))+
theme(legend.text = element_text(size = 16),
legend.title = element_text(size = 20))+
theme(axis.text= element_text(size = 16),
axis.title = element_text(size = 20))


```

```{Mean plots}


meanplot_ordered<-meta_final %>%
mutate(method=factor(method, levels = c("Malaise", "Yellow",  "Fake flower", "Plant", "Wipe", "Netting"))) %>%
arrange(method) %>%
ggerrorplot(x = "method", y = "sr",
desc_stat = "mean_sd",
error.plot = "errorbar",            
add = "mean"                        
) +
theme(legend.text = element_text(size = 16),
legend.title = element_text(size = 20))+
theme(axis.text= element_text(size = 16),
axis.title = element_text(size = 20))+
labs(y = "OTU richness",
x = "Method") +
stat_n_text()

```

```{bar chart}
tax_order<- new_taxa %>% 
  mutate(order = case_when(str_detect(class, "Collembola") ~ "Collembola",
         TRUE ~ as.character(order))) %>%
  mutate(order = case_when(str_detect(subclass, "Acari") ~ "Acari",
         TRUE ~ as.character(order))) %>% 
  filter(perc_id >= 90) %>% 
  select(otu_id, order) %>%
  mutate(otu_id = tolower(otu_id))

rel_otu_num <- new_otu %>% 
  pivot_longer(cols = -sample, names_to = "otu_id", values_to = "reads") %>% 
     inner_join(tax_order, by = "query") %>%
     drop_na(order) %>% 
     mutate(order = factor(order)) %>% 
     filter(reads > 0) %>%
     group_by(sample, order) %>% 
     summarise(n_queries = n_distinct(query), .groups = "drop") %>% 
     group_by(sample) %>% 
     mutate(proportion = n_queries / sum(n_queries)) %>% 
     ungroup()


####SET COLORS; COLORS ARE TAKEN FROM:
#https://stackoverflow.com/questions/57153428/r-plot-color-combinations-that-are-colorblind-accessible#

safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499",
"#44AA99", "#999903", "#882255", "#000000", "#888888", "#0072B2")

###MAKE PICTURE
a_cleaned_90percent<-rel_otu_num %>%
left_join(meta_final) %>%
na.omit() %>%
select(sample, order, type, proportion) %>%
group_by(type) %>%
mutate(sum_time = sum(proportion)) %>%
ungroup() %>%
group_by(type, order) %>%
mutate(rel_abund_time = sum(proportion)) %>%
ungroup() %>%
group_by(type,order) %>%
summarize(rel_abund_time = mean(rel_abund_time/sum_time)) %>%
mutate(order2 = case_when(rel_abund_time < .01 ~ "< 1% rel. number",
TRUE ~ as.character(order))) %>%
group_by(order2,type) %>% summarise(rel_abund_time  = sum(rel_abund_time)) %>%
ggplot() +
geom_col(aes(x = type, y = rel_abund_time, fill = order2), color = "black") +
scale_fill_manual(values = safe_colorblind_palette) +
theme_bw() +
scale_y_continuous(labels = scales::percent_format())+
xlab("Methods") +
ylab("Relative number of OTUs (%)") +
guides(fill=guide_legend(title="Taxa")) +
theme(legend.text = element_text(size = 18),
legend.title = element_text(size = 20))+
theme(axis.text= element_text(size = 18),
axis.title = element_text(size = 20))

###write.csv(a_cleaned_90percent[["data"]], "rel_abund_1percent.csv")

###for 100% species data; some stuff was BLASTed, so we knew what it was

species_check <- new_taxa %>%
mutate(species = case_when(str_detect(species, "Smittia sp. BIOUG21999-A08") ~ "Bryophaenocladius terrestris",
TRUE ~ as.character(species))) %>%
mutate(species = case_when(str_detect(species, "Athalia sp. O86") ~ "Athalia rosae",
TRUE ~ as.character(species))) %>%
filter(perc_id == 100) %>%
select(otu_id, species) %>%
mutate(otu_id = tolower(otu_id))

rel_abund_species_sample <- new_otu %>%
pivot_longer(cols = -sample, names_to = "otu_id", values_to = "reads") %>%
left_join(species_check , by = "otu_id") %>%
drop_na(species) %>%
mutate(species = factor(species)) %>%
group_by(sample, species) %>%
summarise(read_species = sum(reads)) %>%
ungroup() %>%
group_by(sample) %>%
mutate(read_sum = sum(read_species)) %>%
ungroup() %>%
group_by(sample, species) %>%
summarise(rel_abund  = read_species/read_sum)

species_method<-rel_abund_species_sample %>%
left_join(meta) %>%
filter(sample != "AFGr", sample != "AFGrO") %>%
select(sample, species, method, rel_abund) %>%
group_by(method) %>%
mutate(sum_time = sum(rel_abund)) %>%
ungroup() %>%
group_by(method, species) %>%
mutate(rel_abund_time = sum(rel_abund)) %>%
ungroup() %>%
group_by(method,species) %>%
summarize(rel_abund_time = mean(rel_abund_time/sum_time))


###figure with nmds; patchwork

meanplot_ordered + nmds_method / a_cleaned_90percent

```

```{UpSet Plot}

all_otus <- unique(unlist(venn_data_methods_pests))
sets <- names(venn_data_methods_pests)

upset_df <- data.frame(
row.names = all_otus,
sapply(sets, function(s) as.integer(all_otus %in% venn_data_methods_pests[[s]]))
)

###small change

upset_df$"Fake Flower"<-upset_df$Fake.Flower

upset_df$"Yellow pan trap"<-upset_df$Yellow.pan.trap

upset_df<-upset_df %>% select(-Yellow.pan.trap, -Fake.Flower)

###plot

upset(
upset_df,
sets,               # the column names to use as sets
min_size = 7
)


youwaupset<-upsetme + theme(text= element_text(size = 16),title = element_text(size = 16))


